---
title: "Intro to R"
output: pdf_document
---

# Week 1

### Monday

On Monday we did a brief introduction to R. We looked at making functions, variables, and played around with string substitution and printing out a little message.

```{r}
function1 <- function(name, years){
  age <- 365*years
  final <- paste("Hello", name, "you are", age, "days old!")
  print(final) 
}

function2 <- function(name, years){
  age <- 365*years
  sprintf("Hello %s, you are %d days old!", name, age)
}

function1("Ephraim", 24)
function2("Ephraim", 24)


```

```{r}
players <- read.csv("../data/players.csv")
games_details <- read.csv("../data/games_details.csv")
games <- read.csv("../data/games.csv")
teams <- read.csv("../data/teams.csv")
ranking <- read.csv("../data/ranking.csv")
 
```

### Tuesday

Today we'll continue learning R and playing around with the data a little bit.

Let's start by making a function that says something about you depending on which name you put in.

```{r}
teams$mean_arena <- mean(teams$ARENACAPACITY, na.rm = TRUE)
# hist(teams$ARENACAPACITY, breaks=12, col="red")
print(teams[teams$ARENACAPACITY == 0 , "ABBREVIATION"])







```

We learned about tidyverse and ggplot and the powerful syntax they provide.
```{r}

library("ggplot2")
library("tidyverse")

```

We showed how things could be accomplished in both base R and in tidyverse. 

```{r}

teams$mean_size <- mean(teams$ARENACAPACITY, na.rm = TRUE)

```

```{r}
ranking <- ranking %>% separate(STANDINGSDATE, c("year","month", "day"))


```

We played around with arena information and learned about missing values and adding new variables.
```{r}
teams %>% select(NICKNAME, ARENA, ARENACAPACITY) %>% mutate(mean = mean(ARENACAPACITY, na.rm = TRUE))


```

```{r}
teams %>% filter(ARENACAPACITY != 0) %>%
  ggplot(aes(x=ARENACAPACITY)) +
  geom_density(color="black", fill="white", na.rm = TRUE, binwidth=1000)

```

Plot arena size

```{r}
ranking <- ranking %>% filter(month == "03") %>% group_by(TEAM_ID) %>%  mutate(mean_w = mean(W, na.rm = TRUE))
# ranking %>% sort(mean_w, decreasing = TRUE)

combined_games_details <- merge(games_details,games,by="GAME_ID")
```

Now we can look at the mean difference in home vs away points.

```{r}
mean_home <- mean(games$PTS_home, na.rm = TRUE) 
mean_away <- mean(games$PTS_away, na.rm = TRUE)
mean_home - mean_away
```

We clearly see above that the home team does marginally better. This is what many of us might expect. But the result is not that large. Do we know if this could be just by chance? In other words, what is the probability that the mean difference is instead `0`?

In Statistics we have a way of testing this called a t-test. We'll come back to this.

```{r}
ggplot() + 
  geom_density(data = games, aes(x = PTS_home, fill = "home"), alpha = 0.3) +
  geom_density(data = games, aes(x = PTS_away, fill = "away"), alpha = 0.3) +
  scale_colour_manual(name ="Team", values = c("home" = "red", "away" = "blue"), labels=c("home" = "Home Team", "away" = "Away Team")) +
  scale_fill_manual(name ="Team", values = c("home" = "red", "away" = "blue"), labels=c("home" = "Home Team", "away" = "Away Team")) +
  geom_vline(xintercept=mean_home) + geom_vline(xintercept=mean_away) + labs(title="Home vs Away Points",
        x ="Points", y = "Density")
```

We started looking at season stats for players.
```{r}
# games_details %>% group_by(PLAYER_ID) %>% summarise(mean_plus_minus = mean(PLUS_MINUS, na.rm = TRUE)) %>% select(PLAYER_ID, PLAYER_NAME, PLUS_MINUS, mean_plus_minus) %>% arrange(desc(mean_plus_minus))

games_details %>% group_by(PLAYER_ID, PLAYER_NAME) %>% summarise(mean_plus_minus = mean(PLUS_MINUS, na.rm = TRUE), games = sum(!is.na(PLUS_MINUS))) %>% arrange(desc(mean_plus_minus))

```

```{r}
games_details %>% group_by(PLAYER_ID, PLAYER_NAME) %>% select(PLAYER_ID, PLAYER_NAME, PLUS_MINUS) %>% mutate(mean_plus_minus = mean(PLUS_MINUS, na.rm = TRUE), num_games = sum(!is.na(PLUS_MINUS))) %>% drop_na(PLUS_MINUS) %>% select(-PLUS_MINUS) %>% distinct() %>% arrange(desc(num_games)) %>% filter(num_games >= 250)


```

```{r}
games_details %>% group_by(PLAYER_ID, PLAYER_NAME) %>% summarise(mean_plus_minus = mean(PLUS_MINUS, na.rm = TRUE), num_games = sum(!is.na(PLUS_MINUS))) %>% arrange(desc(mean_plus_minus)) %>% filter(num_games >= 350)
```



# Chef Curry

Here we played around with merging data from different dataframes.
```{r}
players = games_details %>% select(-c("TEAM_ID", "TEAM_CITY", "PLAYER_ID", "COMMENT"))
  
games_date = games[,c("GAME_DATE_EST", "GAME_ID", "SEASON")]

# stats = steph.merge(games_date, on="GAME_ID", how="left")

# stats <- steph %>% left_join(games_date, by = c("GAME_ID"))

stats <- left_join(players, games_date, by = c("GAME_ID"))


head(stats)
```

We then learned about how to chain commands together in the tidyverse (or dplyr) syntax using pipes. We learned how to group datasets by variables of interest and perform computations within these groups.

```{r}
seasonal_stats <- stats %>% group_by(SEASON, PLAYER_NAME) %>% summarise(PTS = mean(PTS, na.rm=TRUE)) %>% arrange(SEASON)
seasonal_stats
```

Last weekends homework.

```{r}

select_players <- function (players) {
 seasonal_stats %>% filter(PLAYER_NAME %in% players) %>%
  ggplot(aes(x=factor(SEASON), y=PTS, group=PLAYER_NAME, fill=PLAYER_NAME, colour=PLAYER_NAME)) + geom_line(stat="identity", show.legend = TRUE) + labs(x="Season", y="Points Per Season") + 
    theme( axis.text.x=element_text(angle=90) )
}

```

```{r}
select_players(players=c("Stephen Curry", "LeBron James", "Jayson Tatum", "Kawhi Leonard"))
```


# Curry vs the rest of the league


Here we learned how to summarise (collapse the data) across multiple columns or variables at once. We then plotted performance for the rest of the team against a player of our choice.

```{r}
all_players <- games_details %>% select(PLAYER_NAME, FGM, FG_PCT, FG3_PCT, PTS, FG3M, FG3A, FTM, FT_PCT) %>% group_by(PLAYER_NAME) %>% summarise_all(sum, na.rm = TRUE)

ggplot() + geom_point(data=all_players, aes(x=FGM, y=PTS)) + geom_point(data=all_players[all_players$PLAYER_NAME == "Stephen Curry",], aes(x=FGM, y=PTS), color='red') + labs(x="Field Goals Made (FGM)", y="Points (PTS)", title="Steph Curry vs the league", subtitle ="Points vs Number of Fieldgoals Made" ) + geom_abline(slope = 2, intercept = 0)
```

