---
title: "Intro to R"
output: pdf_document
---

# Week 1

### Monday

On Monday we did a brief introduction to R. We looked at making functions, variables, and played around with string substitution and printing out a little message.

```{r}
function1 <- function(name, years){
  age <- 365*years
  final <- paste("Hello", name, "you are", age, "days old!")
  print(final) 
}

function2 <- function(name, years){
  age <- 365*years
  sprintf("Hello %s, you are %d days old!", name, age)
}

function1("Ephraim", 24)
function2("Ephraim", 24)


```

```{r}
players <- read.csv("../data/players.csv")
games_details <- read.csv("../data/games_details.csv")
games <- read.csv("../data/games.csv")
teams <- read.csv("../data/teams.csv")
ranking <- read.csv("../data/ranking.csv")
 
```

### Tuesday

Today we'll continue learning R and playing around with the data a little bit.

Let's start by making a function that says something about you depending on which name you put in.

```{r}
teams$mean_arena <- mean(teams$ARENACAPACITY, na.rm = TRUE)
# hist(teams$ARENACAPACITY, breaks=12, col="red")
print(teams[teams$ARENACAPACITY == 0 , "ABBREVIATION"])



```

```{r}

library("ggplot2")
library("tidyverse")

```

Next we're going to play around with this. let's create a new column two ways. First using base. And then using dplyr

```{r}

teams$mean_size <- mean(teams$ARENACAPACITY, na.rm = TRUE)

```

```{r}
ranking <- ranking %>% separate(STANDINGSDATE, c("year","month", "day"))


```

```{r}
teams %>% select(NICKNAME, ARENA, ARENACAPACITY) %>% mutate(mean = mean(ARENACAPACITY, na.rm = TRUE))


```

```{r}
teams %>% filter(ARENACAPACITY != 0) %>%
  ggplot(aes(x=ARENACAPACITY)) +
  geom_density(color="black", fill="white", na.rm = TRUE, binwidth=1000)

```

Plot arena size

```{r}
ranking <- ranking %>% filter(month == "03") %>% group_by(TEAM_ID) %>%  mutate(mean_w = mean(W, na.rm = TRUE))
# ranking %>% sort(mean_w, decreasing = TRUE)

combined_games_details <- merge(games_details,games,by="GAME_ID")
```

Now we can look at the mean difference in home vs away points.

```{r}
mean_home <- mean(games$PTS_home, na.rm = TRUE) 
mean_away <- mean(games$PTS_away, na.rm = TRUE)
mean_home - mean_away
```

We clearly see above that the home team does marginally better. This is what many of us might expect. But the result is not that large. Do we know if this could be just by chance? In otherwords, what is the probability that the mean difference is instead `0`?

In Statistics we have a way of testing this called a t-test.

```{r}
ggplot() + 
  geom_density(data = games, aes(x = PTS_home, fill = "home"), alpha = 0.3) +
  geom_density(data = games, aes(x = PTS_away, fill = "away"), alpha = 0.3) +
  scale_colour_manual(name ="Team", values = c("home" = "red", "away" = "blue"), labels=c("home" = "Home Team", "away" = "Away Team")) +
  scale_fill_manual(name ="Team", values = c("home" = "red", "away" = "blue"), labels=c("home" = "Home Team", "away" = "Away Team")) +
  geom_vline(xintercept=mean_home) + geom_vline(xintercept=mean_away) + labs(title="Home vs Away Points",
        x ="Points", y = "Density")
```

```{r}
# games_details %>% group_by(PLAYER_ID) %>% summarise(mean_plus_minus = mean(PLUS_MINUS, na.rm = TRUE)) %>% select(PLAYER_ID, PLAYER_NAME, PLUS_MINUS, mean_plus_minus) %>% arrange(desc(mean_plus_minus))

games_details %>% group_by(PLAYER_ID, PLAYER_NAME) %>% summarise(mean_plus_minus = mean(PLUS_MINUS, na.rm = TRUE), games = sum(!is.na(PLUS_MINUS))) %>% arrange(desc(mean_plus_minus))

```

```{r}
games_details %>% group_by(PLAYER_ID, PLAYER_NAME) %>% select(PLAYER_ID, PLAYER_NAME, PLUS_MINUS) %>% mutate(mean_plus_minus = mean(PLUS_MINUS, na.rm = TRUE), num_games = sum(!is.na(PLUS_MINUS))) %>% drop_na(PLUS_MINUS) %>% select(-PLUS_MINUS) %>% distinct() %>% arrange(desc(num_games)) %>% filter(num_games >= 250)


```

```{r}
games_details %>% group_by(PLAYER_ID, PLAYER_NAME) %>% summarise(mean_plus_minus = mean(PLUS_MINUS, na.rm = TRUE), num_games = sum(!is.na(PLUS_MINUS))) %>% arrange(desc(mean_plus_minus)) %>% filter(num_games >= 350)
```

```{r}
ggplot() +
  geom_density(data = games, aes(x = PTS_home, fill = "Blue"), alpha=0.3) + 
    geom_density(data = games, aes(x = PTS_away, fill = "Red"), alpha = 0.3) 

               
                # geom_density(data = games, aes(x = PTS_home, fill = "r"), alpha = 0.3) +

```

# Chef Curry

```{r}
players = games_details %>% select(-c("TEAM_ID", "TEAM_CITY", "PLAYER_ID", "COMMENT"))
  
games_date = games[,c("GAME_DATE_EST", "GAME_ID", "SEASON")]

# stats = steph.merge(games_date, on="GAME_ID", how="left")

# stats <- steph %>% left_join(games_date, by = c("GAME_ID"))

stats <- left_join(players, games_date, by = c("GAME_ID"))


head(stats)
```

```{r}
# seasonal_stats = stats.groupby("SEASON").sum()/stats.groupby("SEASON").count()
seasonal_stats <- stats %>% group_by(SEASON, PLAYER_NAME) %>% summarise(PTS = mean(PTS, na.rm=TRUE)) %>% arrange(SEASON) %>% ungroup() %>% group_by(PLAYER_NAME) %>% filter(any(PTS > 25))
seasonal_stats
```






```{r}

select_players <- function (players) {
 seasonal_stats %>% filter(PLAYER_NAME %in% players) %>%
  ggplot(aes(x=factor(SEASON), y=PTS, group=PLAYER_NAME, fill=PLAYER_NAME, colour=PLAYER_NAME)) + geom_line(stat="identity", show.legend = TRUE) + labs(x="Season", y="Points Per Season") + theme(
    axis.text.x=element_text(angle=90),
  )
  
}

```


```{r}
select_players(players=c("Stephen Curry", "LeBron James", "Jayson Tatum"))
```




# Curry vs the rest of the league

```{r}
all_players <- games_details %>% select(PLAYER_NAME, FGM, FG_PCT, FG3_PCT, PTS, FG3M, FG3A, FTM, FT_PCT) %>% group_by(PLAYER_NAME) %>% summarise_all(sum, na.rm = TRUE)

ggplot(all_players, aes(x=FGM, y=PTS)) + geom_point() + geom_point(data=all_players[all_players$PLAYER_NAME == "Stephen Curry",], aes(x=FGM, y=PTS), color='red') + labs(x="Points (PTS)", y="Field Goals Made (FGM)", title="Steph Curry vs the league", subtitle ="Points vs Number of Fieldgoals Made" )
```



```{r}
# fig, axes = plt.subplots(2,2, figsize=(15,10))
# fig.suptitle("Stephen Curry vs Other Players Since 2003", fontsize=20)
# 
# sns.scatterplot(all_players["FGM"], all_players["PTS"], ax = axes[0][0])
# sns.scatterplot(x="FGM", y="PTS", data=all_players[all_players.index == "Stephen Curry"], ax = axes[0][0])
# axes[0][0].title.set_text('Field Goals Made and Points')
# 
# sns.scatterplot(all_players["FGM"], all_players["FG_PCT"], ax = axes[0][1])
# sns.scatterplot(x="FGM", y="FG_PCT", data=all_players[all_players.index == "Stephen Curry"], ax = axes[0][1])
# axes[0][1].title.set_text('Field Goals Made and Percantages')
# 
# sns.scatterplot(all_players["FG3M"], all_players["FG3_PCT"], ax = axes[1][0])
# sns.scatterplot(x="FG3M", y="FG3_PCT", data=all_players[all_players.index == "Stephen Curry"], ax = axes[1][0])
# axes[1][0].title.set_text('Three Point Made and Attempts')
# 
# sns.scatterplot(all_players["FTM"], all_players["FT_PCT"], ax = axes[1][1])
# sns.scatterplot(x="FTM", y="FT_PCT", data=all_players[all_players.index == "Stephen Curry"], ax = axes[1][1])
# axes[1][1].title.set_text('Free Throw Made and Percantages')
# 
# fig.tight_layout()
```
