---
title: "Intro to R"
output: pdf_document
---

# Week 1

### Monday

On Monday we did a brief introduction to R. We looked at making functions, variables, and played around with string substitution and printing out a little message.

```{r}
function1 <- function(name, years){
  age <- 365*years
  final <- paste("Hello", name, "you are", age, "days old!")
  print(final) 
}

function2 <- function(name, years){
  age <- 365*years
  sprintf("Hello %s, you are %d days old!", name, age)
}

function1("Ephraim", 24)
function2("Ephraim", 24)


```

```{r}
players <- read.csv("../data/players.csv")
games_details <- read.csv("../data/games_details.csv")
games <- read.csv("../data/games.csv")
teams <- read.csv("../data/teams.csv")
ranking <- read.csv("../data/ranking.csv")
 
```

### Tuesday

Today we'll continue learning R and playing around with the data a little bit.


Let's start by making a function that says something about you dending on which name you put in.




```{r}
library("ggplot2")
library("tidyverse")

```

Next we're going to play around with this. let's create a new column two ways. First using base. And then using dplyr

```{r}

teams$mean_size <- mean(teams$ARENACAPACITY, na.rm = TRUE)
teams %>% select(NICKNAME, ARENA, ARENACAPACITY) %>% mutate(mean = mean(arenacapacity, na.rm = TRUE))


```

```{r}
names(cbps_hh_ls1 %>%
  select(contains("read")))

```


```{r}
teams %>% filter(ARENACAPACITY != 0) %>%
  ggplot(aes(x=ARENACAPACITY)) +
  geom_histogram(color="black", fill="white", na.rm = TRUE, binwidth=1000)

```


Plot arena size

